On Devon Allen's Disqualification at the 2022 Track and Field World Championship

### 2023-04-10 Manuscript/manuscript.tex
Added few references, moved tables to after end document

1)
After coming in third in the 110 meter hurdles at the United States Track and 
Field Championships in Eugene Oregon at the end of June, Allen was expected to 
be a contender for the World Championships, which were scheduled for two months 
later, also in Eugene. At the World Championships, Allen made it through the 
heats and semifinals to reach the finals while posting reaction times of 0.123
in the heats and 0.101 in the semifinal.  Allen, who attended the University
of Oregon (Location of US World Championship and the World Championship) was
considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
in early June; only 0.04 seconds away from the world record \citep{Preview}.
Thus his disqualification by reacting in 0.099 seconds instead of 0.01 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their Youtube
channel that show the disqualification and can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers when
they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting another two years until the next World Track and Field Championship.

2)
Outside the blog posts from LetsRun.com, there have not been any published
papers regarding Devon Allen's disqualification in particular.  This is not
particularly surprising given that his disqualification occurred only last August.
Letsrun is a website that is part message board and part news.  The message board,
which functions similarly to Reddit is very active during important running
events, which include during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as ``Was Devon Allen
Screwed? There's At Least a 99\% That He Was" and ``The Data Keeps Pouring In and
It Continues to Look Bad For World Athletics and Great For Devon Allen".  Robert
Johnson, who created and runs the Letsrun website argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, statistics such as median reaction time, and compared the reaction
times at the 2022 Track and Field World Championship.

3)
In 2009, the International Association of Athletics Federations (Also referred to
as IAAF or World Athletics), which is the governing board for international running
competitions, commissioned a study into changing the reaction time barrier.
The paper analyzed sprinters reaction times and the biological processes 
that went into reacting to the starting gun and concluded 
that athletes can react in as low as 0.08 seconds.  One study by Pain looked at
reaction times for nine male athletes and analyzed their reaction times.  The
fastest athlete of the nine, had on average (even with their two fastest time 
removed) a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}.
to IAAF that the disqualification barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body movement.
This could provide a more accurate representation of reaction time, and thus the
possibility of a false disqualification would be lowered \citep{komi2009iaaf}.  
A follow-up study from 2012 by Pilianidis, Mantzouranis, and Kasabalis found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that for reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}.  This will
be important when considering the results in section \ref{sec:Results}.

Despite the study, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.  It is confusing
as to why IAAF would commission a study only to disregard the results and not
make the necessary changes.  It is unknown how many athletes since 2009 could
have potentially benefited from this, but it is probable that there are others
besides Allen who were disqualified.
%Reaction time barrier has been examined before: cite reviews
%Then go into specifics and talk about people who wanted to change barrier
%General reviews are important
%People have studied the biological processes, cite a few references
%How barrier is picked by regulator
%Read abstract and into and summarize into about one sentence what each paper contributes
%1 and 2 can be merged but 3 need to stand alone

4)
One of the first objectives of the paper is to try to model the data, which is
the reaction times of every non-disqualification since the 1999 World Championships. 
When determining the most effective model to use, several variations were
explored across the 4 different data sets: a linear model without a venue effect,
linear model with venue effect, a gamma model without a venue effect, a gamma model
with a venue effect, a gamma model without a venue effect and link set to log instead
of the default which is Gaussian.  The R package lme4 was very helpful with this
as it supported the Generalized Linear Mixed Model (GLMM) which allowed 
linear models with various distributions to be modeled\citep{Rpkg:lme4}.  The
model we use is a liner model with a fixed effect and random effect and follows
the gamma distribution. The fixed effect comes in the form of an intercept, and the
random effect will be called the venue effect and is reflective of how significant
the changes in the reaction time is for each venue, i.e. the venue effect is 
responsible for the variation between years.

Another objective of this paper is to compare data for athletes who have competed
at additional competitions to see how their reaction times compare from previous
competitions to this one. First we can look at the reaction times of athletes who competed at the
World Track and Field Championship in both 2019 and 2022 and compare the times,
and secondly we can look at how athletes performed compared to how they did at
their national/qualifiers meet.  We would expect that athletes of this caliber,
who are world-class athletes, to not improve significantly from one year to the
next.  In order to reduce any chances of athletes improving reaction time, we
can also look at data from qualifying meets.  Almost every athlete had to qualify 
by proving they are elite at the national level, usually by coming in at least
fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August.  Thus, it should be expected that 
elite athletes such as these will not get reasonably faster in such a short time
span. The equipment used to measure reaction time cannot be guaranteed to be
the same from national to international competition, but the time span between
meets has been reduced considerably.
%Simplify contribution to one paragraph: collect data and formal statistical
%analysis such as model used
%Move a lot of this to methods section
5)
The rest of this paper is as follows: Section~\ref{sec:Data} describes how data was
collected and begins to detail the GLMM model which is further developed in the
Section~\ref{sec:Methods}.  The signed-rank test analysis for the same athletes is
also included in the Section~\ref{sec:Methods}.  The Section~\ref{sec:Results}
section provides the probabilities from the methods section and the paper ends 
by questioning the reaction time barrier.

@@

Data was copied from the World Athletics website and pasted into an Excel
spreadsheet. The data covers the men's 110 meter hurdles and the women's 100
meter hurdles from 1999 to 2022.  The variable of prediction interest is reaction
time, and the predicting variables are year, stage of competition (heats, 
semifinal, final), total time, and gender.  The data can be found in the appendix:.  
Factors such as name, country, and identification number were
not considered important for the purposes of this research besides the times for
Devon Allen and the other United States athletes mentioned above.  The variable "Stage"
refers to whether the observation occurred during the "heats" (Preliminary round) which
is denoted by "H" in the data, "S" for "semifinals" (Of which there are either 2 or 3 
heats each year), and "F" for finals. The variable "Gender" has levels "M" and "F" 
which stand for "male" and "female" respectively.  The variable "Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly).  There is no statistical signficance to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary Heat of 2022 and Batch
111 corresponds to the 1999 finals). The data that looked at in the paper
is best visualized using a box plot. Figure~\ref{fig:DataBoxPlot}.  Included are
box plots of the reaction times of the men's semi finals and finals pooled
together and just the finals.
%Fix quotations
%Dont need the titles will add captions, use width =5, height = 3 in aspect ratio
\begin{figure}[h]
  \centering 
  \includegraphics[width=\textwidth,height=4in]{DataBoxPlot}
  \caption{Boxplot of semifinals and finals data}\label{fig:DataBoxPlot}
\end{figure}
\jy{The plot was stretched. Don't. Get their sizes right before generating them.}

\jy{Remove title from the figures. Make the figure captions self-contained.}
\begin{figure}[h]
  \centering 
  \includegraphics{BoxplotFinals}
  \label{fig:BoxplotFinals}
\end{figure}

>>>>

%1
After coming in third in the 110 meter hurdles at the United States Track and 
Field Championships in Eugene Oregon at the end of June, Allen was expected to 
be a contender for the World Championships, which were scheduled for two months 
later, also in Eugene. At the World Championships, Allen made it through the 
heats and semifinals to reach the finals while posting reaction times of 0.123
in the heats and 0.101 in the semifinal.  Allen, who attended the University
of Oregon (Location of US World Championship and the World Championship) was
considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
in early June; only 0.04 seconds away from the world record \citep{Preview}.
Thus his disqualification by reacting in 0.099 seconds instead of 0.01 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their Youtube
channel that show the disqualification and can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers when
they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting another two years until the next World Track and Field Championship.

%2
Outside the blog posts from LetsRun.com, there have not been any published
papers regarding Devon Allen's disqualification in particular.  This is not
particularly surprising given that his disqualification occurred only last August.
Letsrun is a website that is part message board and part news.  The message board,
which functions similarly to Reddit is very active during important running
events, which include during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as ``Was Devon Allen
Screwed? There's At Least a 99\% That He Was" and ``The Data Keeps Pouring In and
It Continues to Look Bad For World Athletics and Great For Devon Allen".  Robert
Johnson, who created and runs the Letsrun website argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, statistics such as median reaction time, and compared the reaction
times at the 2022 Track and Field World Championship.

%3
In 2009, the International Association of Athletics Federations (Also referred to
as IAAF or World Athletics), which is the governing board for international running
competitions, commissioned a study into changing the reaction time barrier.
The paper analyzed sprinters reaction times and the biological processes 
that went into reacting to the starting gun and concluded 
that athletes can react in as low as 0.08 seconds.  One study by Pain looked at
reaction times for nine male athletes and analyzed their reaction times.  The
fastest athlete of the nine, had on average (even with their two fastest time 
removed) a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}.
to IAAF that the disqualification barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body movement.
This could provide a more accurate representation of reaction time, and thus the
possibility of a false disqualification would be lowered \citep{komi2009iaaf}.  
A follow-up study from 2012 by Pilianidis, Mantzouranis, and Kasabalis found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that for reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}.  However, not
all studies have been in favor of a decrease in the reaction time barrier.
Brosnan in 2016 argues for increasing the reaction time barrier to 0.115 seconds,
and cited evidence from European Championships and World Championships in his
article \citep{brosnan2017effects}.  A study looking at the 2008 Olympics
likewise supported the conclusion that athletes could not react in as little as
0.1 seconds \citep{lipps2011implications}.
%About timing
It has been concluded that

%Reaction time barrier has been examined before: cite reviews
%Then go into specifics and talk about people who wanted to change barrier
%General reviews are important
%People have studied the biological processes, cite a few references
%How barrier is picked by regulator
%Read abstract and into and summarize into about one sentence what each paper contributes
%1 and 2 can be merged but 3 need to stand alone

%4
One of the first objectives of the paper is to try to model the data, which is
the reaction times of every non-disqualification since the 1999 World Championships. 
When determining the most effective model to use, several variations were
explored across the 4 different data sets: a linear model without a venue effect,
linear model with venue effect, a gamma model without a venue effect, a gamma model
with a venue effect, a gamma model without a venue effect and link set to log instead
of the default which is Gaussian.  The R package lme4 was very helpful with this
as it supported the Generalized Linear Mixed Model (GLMM) which allowed 
linear models with various distributions to be modeled\citep{Rpkg:lme4}.  The
model we use is a liner model with a fixed effect and random effect and follows
the gamma distribution. The fixed effect comes in the form of an intercept, and the
random effect will be called the venue effect and is reflective of how significant
the changes in the reaction time is for each venue, i.e. the venue effect is 
responsible for the variation between years. 
Another objective of this paper is to compare data for athletes who have competed
at additional competitions to see how their reaction times compare from previous
competitions to this one. First we can look at the reaction times of athletes who competed at the
World Track and Field Championship in both 2019 and 2022 and compare the times,
and secondly we can look at how athletes performed compared to how they did at
their national/qualifiers meet.  We would expect that athletes of this caliber,
who are world-class athletes, to not improve significantly from one year to the
next.  In order to reduce any chances of athletes improving reaction time, we
can also look at data from qualifying meets.  Almost every athlete had to qualify 
by proving they are elite at the national level, usually by coming in at least
fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August.  Thus, it should be expected that 
elite athletes such as these will not get reasonably faster in such a short time
span. The equipment used to measure reaction time cannot be guaranteed to be
the same from national to international competition, but the time span between
meets has been reduced considerably.
%Simplify contribution to one paragraph: collect data and formal statistical
%analysis such as model used
%Move a lot of this to methods section

%5
The rest of this paper is as follows: Section~\ref{sec:Data} describes how data was
collected and begins to detail the GLMM model which is further developed in the
Section~\ref{sec:Methods}.  The signed-rank test analysis for the same athletes is
also included in the Section~\ref{sec:Methods}.  The Section~\ref{sec:Results}
section provides the probabilities from the methods section and the paper ends 
by questioning the reaction time barrier.

@@

Data was copied from the World Athletics website and pasted into an Excel
spreadsheet. The data covers the men's 110 meter hurdles and the women's 100
meter hurdles from 1999 to 2022.  The variable of prediction interest is reaction
time, and the predicting variables are year, stage of competition (heats, 
semifinal, final), total time, and gender.  The data can be found in the appendix:.  
Factors such as name, country, and identification number were
not considered important for the purposes of this research besides the times for
Devon Allen and the other United States athletes mentioned above.  The variable "Stage"
refers to whether the observation occurred during the "heats" (Preliminary round) which
is denoted by "H" in the data, "S" for "semifinals" (Of which there are either 2 or 3 
heats each year), and "F" for finals. The variable "Gender" has levels "M" and "F" 
which stand for "male" and "female" respectively.  The variable "Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly).  There is no statistical signficance to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary Heat of 2022 and Batch
111 corresponds to the 1999 finals). The data that looked at in the paper
is best visualized using a box plot as show in Figure~\ref{fig:Boxplots}.  Included are
box plots of the reaction times of the men's semi finals and finals pooled
together and just the finals.  It is clear that the 2022 boxplot for both the
men's finals and semifinals data is lower (indicating a faster reaction time)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
Brosnan found reaction times of 0.156 seconds for men when looking at data from
1999 to 2014 \citep{brosnan2017effects}.
%Fix quotations
%Dont need the titles will add captions, use width =5, height = 3 in aspect ratio

\begin{figure}[h]
  \centering
  \includegraphics{Finals_Pooled_Boxplot}
  \label{fig:Boxplots}
  \caption{The reaction times from 2019 to 2022 for the men's final (top) and
  the men's semifinals and finals (bottom)}
\end{figure}
